# Regression and model validation

> ### Instructions
> Describe the work you have done this week and summarize your learning.
>
> - Describe your work and results clearly. 
> - Assume the reader has an introductory course level understanding of > writing and reading R code as well as statistical methods.
> - Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

## 2.0 Packages

Let's first load the required packages.

```{r}
pacman::p_load(tidyverse, GGally)
```

## 2.1 Reading data from the web

Let's begin by exploring our dataset. Click here for [metadata](http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-meta.txt).

```{r}
lrn14 <- read.table("http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-data.txt", sep = "\t", header = TRUE)
dim(lrn14)
# str(lrn14)
```

### Observations from the data:

- Looks like the `lrn14` dataset has `r nrow(lrn14)` rows and `r ncol(lrn14)` columns.
- Most columns appear to be intergers, whereas only `gender` is character.


## 2.2 Scaling variables

> ### Instructions
> - Execute the example codes to see how vectorized division works
> - Use vector division to create a new column `attitude` in the `lrn14` data frame, where each observation of `Attitude` is scaled back to the original scale of the questions, by dividing it with the number of questions.
> 
> Hint:
> - Assign 'Attitude divided by 10' to the new column 'attitude.

```{r}
lrn14 <- lrn14 %>%
  mutate(attitude = Attitude / 10)
hist(lrn14$attitude)

```



## 2.3 Combining variables
## 2.4 Selecting columns

> ### Instructions
> - Access the **dplyr** library
> - Create object `keep_columns`
> - Use `select()` (possibly together with `one_of()`) to create a new data frame `learning2014` with the columns named in `keep_columns`.
> - Look at the structure of the new dataset
> 
> Hint:
> - See the previous exercise or the data wrangling cheatsheet for help on how to select columns

```{r}

```


## 2.5 Modifying column names

## 2.6 Excluding observations

## 2.7 Visualizations with ggplot2

## 2.8 Exploring a data frame

## 2.9 Simple regression

## 2.10 Multiple regression

## 2.11 Graphical model validation

## 2.12 Making predictions
